Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2024-12-17 index.qmd
Minor changes

In data analysis, experienced researchers often rely on their prior knowledge or domain expertise to quickly assess whether results align with their expectations. When a result falls outside of this interval, it prompts the researchers to investigate backwards on the data quality, the analysis steps, or the assumptions made during the analysis process. This mental process of where to diagnose unexpected outcomes is often difficult to trace and discuss in publications. As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned, achieved either by refining the analysis or updating the expectations based on statistical evidence [@grolemund_cognitive_2014]. These missing pieces of information provides little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes. Similarly, when researchers with different background knowledge view results that they find to be unexpected, it becomes unclear whether discrepancies arise from differing expectations or from the use of statistical techniques.

@@

The skill of diagnosing unexpected data analysis results is not one that has received significant attention in the statistics literature. While the concept of diagnosis is often embedded in model checking or data visualization techniques, systematic approaches to identifying the root cause of an unexpected analysis result are typically not presented [@peng2022perspective]. @peng_diagnosing_2021 proposed a series of exercises for training students in data analysis to diagnose different kinds of analysis problems such as coding errors or outliers. They provide a systematic approach involving working backwards from the analysis result to identify potential causes. There are parallels here to the concept of debugging and testing in software engineering [@donoghue2021teaching]. For example, @li2019towards found that experienced engineers were generally able to identify problems in code faster than novices, and that the ability to debug code required knowledge that cut across different domains.

@@

Expectations represent our understanding of certain aspects of the analysis and the data, independent of the results of the analysis itself. When observed outcomes deviate from these expectations, analysts often revisit the analysis process to identify potential issues, refine methods, or revise assumptions. Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, making it harder for less experienced researchers to learn and master these skills.

@@

Consider a 30-day step count experiment in public health. Subjects are instructed to walk at least 8,000 steps each day, with an expected average of 9,000 steps, tracked by a step counter app. After 30 days, we review the data and examine the number of steps taken each day. With data of this nature, we may expect there to be occasional "low" days due to factors such as forgetting to wear their watch or unfavorable weather conditions limiting outdoor activities.  We may also expect "high" days recorded after an outdoor hike or intense workout. Given the requirements of the study, we expect that for a given subject 
the average step count would be between $[8500, 9500]$.

To diagnose potential reasons why this outcome expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data. For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than two-thirds exceed 10,000 steps, this could indicate an excess of low-count or high-count days. Similarly, we may may expect the standard deviation of the step count not to be overly large. These considerations yield the following three analysis validation checks that *fails* when:

@@

From the simulated data, the accuracy branch refers to a set of checks' ability to accurately detect unexpected outcomes while minimizing false positives and false negatives. While a false positive can raise caution or skepticism on the data, the presence of both false positives and false negatives suggest that the checks are overly sensitive or lack sensitivity to unexpected outcomes.

@@

While checks may score high on accuracy, they may be less effective at explaining the various reasons behind unexpected results. This could happen if, for example, a set of checks are all tangentially related to the cause of the unexpected results, but none addresses the root cause. It may also occur if the checks are highly correlated with one another, leading to redundancy.

@@

In the study of the health effects of outdoor air pollution, one area of interest is the association between short-term, day-to-day changes in particulate matter air pollution and daily mortality counts. Substantial work has been done to study this question and to date, there appears to be strong evidence of an association between particulate matter less than 10 $\mu$g/m$^3$ in aerodynamic diameter (PM10) and daily mortality from all non-accidental causes [@samet2000fine]. For our second example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied.

@@

| the check fails if ... | 
|--------------| 
| Sample size less than or equal to 200 |
| Sample size less than or equal to 400 |
| Sample size less than or equal to 600 |
| Sample size less than or equal to 800 |
| Mortality-PM10 correlation less than -0.03 |
| Mortality-PM10 correlation less than -0.04 |
| Mortality-PM10 correlation less than -0.05 |
| Mortality-PM10 correlation less than -0.06 |
| Mortality-temperature correlation greater than -0.3 | 
| Mortality-temperature correlation greater than -0.35 | 
| Mortality-temperature correlation greater than -0.4 | 
| Mortality-temperature correlation greater than -0.45 | 
| Outlier(s) are presented in the variable PM10 | 
| Outlier(s) are presented in the variable mortality |

@@

To ensure a reasonable likeness to data that might be used in such an analysis, we use characteristics of the observed dataset to refine our simulations. AIC is used to determine the best distribution fit for each variable with the QQ-plot presented in @fig-dist-fit to evaluate the fit. AIC suggests a negative binomial distribution for mortality, a beta distribution for PM10 (multiple by 100 to recover the original scale), and a Weibull distribution for temperature. To include the potential effect of outliers, we add a single outlier to the data for both the mortality and PM10 variables.

@@

In this paper we have developed an approach to using analysis validation checks to externalize the assumptions about the data and analysis tools made during the data analysis process. These checks can serve as a useful summary of the analyst's thought process and can describe how characteristics of the data may lead to unexpected outcomes. Using logic regression, we can develop a graphical summary of the analysis validation checks as well as use the logic regression fitting process to choose the optimal set of checks. The logic regression model can also be used to develop summaries of the precision and recall of the collection of analysis validation checks in predicting the likelihood of an unexpected outcome. We demonstrated our method on an example relating daily mortality to outdoor air pollution data.

@@

Systematically generating simulated data is a key component of our approach. In the PM10 example, the inverse transform method is used to preserve the correlation structure among mortality, PM10, and temperature. However, simulation can become complex when additional restrictions are imposed. In such cases, techniques like the acceptance-reject method or permutation may be used to generate the data.

@@

>>>>

In data analysis, experienced researchers often rely on their prior knowledge or domain expertise to quickly assess whether results align with their expectations. When a result falls outside of this expectation, it prompts the researchers to investigate backwards on the data quality, the analysis steps, or the assumptions made during the analysis process. This mental process of where to diagnose unexpected outcomes is often difficult to trace and document in publications. As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned, achieved either by refining the analysis or updating the expectations based on statistical evidence [@grolemund_cognitive_2014]. These missing pieces of information provide little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes. Similarly, when researchers with different background knowledge view results that they find to be unexpected, it becomes unclear whether discrepancies arise from differing expectations or from the use of statistical techniques.

@@

The skill of diagnosing unexpected data analysis results is not one that has received significant attention in the statistics literature. While the concept of diagnosis is often embedded in model checking or data visualization techniques, systematic approaches to identifying the root cause of a general unexpected analysis result are typically not presented [@peng2022perspective]. Furthermore, if interesting information is discovered through model checking or data visualization, there is no formal way to document such discoveries for the next analyst to consider. @peng_diagnosing_2021 proposed a series of exercises for training students in data analysis to diagnose different kinds of analysis problems such as coding errors or outliers. They provide a systematic approach involving working backwards from the analysis result to identify potential causes. There are parallels here to the concept of debugging and testing in software engineering [@donoghue2021teaching]. For example, @li2019towards found that experienced engineers were generally able to identify problems in code faster than novices, and that the ability to debug code required knowledge that cut across different domains.

@@

Expectations represent our understanding of certain aspects of the analysis and the data, independent of the results of the analysis itself. When observed outcomes deviate from these expectations, analysts often revisit the analysis process to identify potential issues, refine methods, or revise assumptions. Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, limiting transparency and making it harder for less experienced researchers to learn and master these skills.

@@

Consider a 30-day step count experiment in a public health setting. Subjects are instructed to walk at least 8,000 steps each day, with an expected average of 9,000 steps, tracked by a step counter app. After 30 days, we review the data and examine the number of steps taken each day. With data of this nature, we may expect there to be occasional "low" days due to factors such as forgetting to wear their watch or unfavorable weather conditions limiting outdoor activities.  We may also expect "high" days recorded after an outdoor hike or intense workout. Given the requirements of the study, we expect that for a given subject 
the average step count would be between $[8500, 9500]$.

To diagnose potential reasons why this outcome expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data. For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than a third exceed 10,000 steps, this could indicate an excess of low-count or high-count days. Similarly, we may may expect the standard deviation of the step count not to be overly large. These considerations yield the following three analysis validation checks that *fail* when:

@@

From the simulated data, the accuracy branch in @fig-metric-calc refers to a set of checks' ability to accurately detect unexpected outcomes while minimizing false positives and false negatives. While a false positive can raise caution or skepticism about the data, the presence of both false positives and false negatives suggests that the checks are overly sensitive or lack sensitivity to unexpected outcomes, respectively.

@@

While checks may score high on predictive accuracy, they may be less effective at explaining the various reasons behind unexpected results. This could happen if, for example, a set of checks are all tangentially related to the cause of the unexpected results, but none addresses the root cause. It may also occur if the checks are highly correlated with one another, leading to redundancy.

@@

In the study of the health effects of outdoor air pollution, one area of interest is the association between short-term, day-to-day changes in particulate matter air pollution and daily mortality counts. Substantial work has been done to study this question and to date, there appears to be strong evidence of an association between particulate matter less than 10 $\mu$g/m$^3$ in aerodynamic diameter (PM10) and daily mortality from all non-accidental causes [@samet2000fine]. For our second example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied. The dataset is from New York City, and contains daily PM10, mortality, and temperature values from 1992--2000.

@@

| the check fails if ... | 
|--------------| 
| Sample size less than or equal to 200 |
| Sample size less than or equal to 400 |
| Sample size less than or equal to 600 |
| Sample size less than or equal to 800 |
| Mortality-PM10 correlation less than $-$0.03 |
| Mortality-PM10 correlation less than $-$0.04 |
| Mortality-PM10 correlation less than $-$0.05 |
| Mortality-PM10 correlation less than $-$0.06 |
| Mortality-temperature correlation greater than $-$0.3 | 
| Mortality-temperature correlation greater than $-$0.35 | 
| Mortality-temperature correlation greater than $-$0.4 | 
| Mortality-temperature correlation greater than $-$0.45 | 
| Outlier(s) are presented in the variable PM10 | 
| Outlier(s) are presented in the variable mortality |

@@

To ensure a reasonable likeness to data that might be used in such an analysis, we use characteristics of the observed New York City dataset to refine our simulations. AIC is used to determine the best distribution fit for each variable with the QQ-plot presented in @fig-dist-fit to evaluate the fit. AIC suggests a negative binomial distribution for mortality, a beta distribution for PM10 (multiple by 100 to recover the original scale), and a Weibull distribution for temperature. To include the potential effect of outliers, we add a single outlier to the data for both the mortality and PM10 variables.

@@

In this paper we have developed an approach to using analysis validation checks to externalize the assumptions about the data and analysis tools made during the data analysis process. These checks can serve as a useful summary of the analyst's thought process and can describe how characteristics of the data may lead to unexpected outcomes. Using logic regression, we can develop a graphical summary of the analysis validation checks as well as use the logic regression fitting process to choose the optimal set of checks. The logic regression model can also be used to develop summaries of the precision and recall of the collection of analysis validation checks in predicting the likelihood of an unexpected outcome. We demonstrated our method on an example relating daily mortality to outdoor air pollution data. The results from that example could be used to inform future analyses of air pollution and health data, perhaps in other cities or locations.

@@

Systematically generating realistic simulated data is a key component of our approach and is deserving of further consideration. In the PM10 example, the inverse transform method is used to preserve the correlation structure among mortality, PM10, and temperature. However, simulation can become complex when additional restrictions are imposed or when a greater range of scenarios is desired. In such cases, techniques like the acceptance-reject method or permutation may be used to generate the data.

@@

Communicating the process of data analysis is a key element to providing transparency, improving reproducibility, and building trust with a range of audiences. The analysis validation checks described here provide a general way to encode the assumptions that an data analyst makes about the data and statistical tools applied to the data. These assumptions can be studied or challenged, depending on the specific analyst's perspective, and can serve as a roadmap for diagnosing unexpected results.
